---
title: "Lab 1 Practice"
author: "Chelsea Sanford"
format: 
    html:
      code-fold: true
      toc: true
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r}
library(tidyverse)
library(gapminder)

view(gapminder)
```

1.  In your github account, create a new repository called "lab1_practice", then clone it to your laptop

In the new R project:

1.  Create a new Quarto doc. If you like, set up the header for code folding. Set echo, message, and warning like we did in lab.
2.  Attach tidyverse and the gapminder package (you might need to install gapminder)
3.  For each year and continent, summarize the average per-capita GDP across all countries in the continent. hint: don't just average the country-level values - need to consider country populations!
4.  Plot the continent_level GDP values in a scatter plot. Bonus, plot the country level values as well with a low alpha to fade the into the background.
5.  Redo the scatter plot but this time, log transform the GDP data (either int he dataframe using 'mutate' or on the plot with 'scale_y"log10()')

based on the plot, choose either regular or log transformed data, and use a linear regression to estimate annual change in per capita GDP **OR** log (per capita GDP) accounting for continent.

-   Write a sentence or two describing the trends in the plots.
-   Write a sentence or two to explain the coefficients in the linear model, including statistical sig.
    -   Which continent is the "reference" continent? why mihgt that be?
    -   What do the coeffecients on each continent represent? Do these make intuitive sense?
-   does a linear model seem like a good way to approach these data? why or why not?

when done, render your quarto doc, then stage, commit, and push back to Github.

# 3

```{r}
percapgdp <- gapminder %>% 
  group_by(continent, year) %>% 
  summarize(percapgdp = mean (gdpPercap))

percapgdp
```

# 4

```{r}
gdp_plot <- ggplot(data = percapgdp, aes(x = year, y = percapgdp, color = continent)) +
  geom_point() +
  labs(x = 'Year', y = 'Per Capita GDP', color = 'Continent', title = 'Continent-Level Per Capita GDP') +
  theme_minimal()

gdp_plot
```

# 5

```{r}
gdp_plot_log <- ggplot(data = percapgdp, aes(x = year, y = percapgdp, color = continent)) +
  geom_point() +
  labs(x = 'Year', y = 'Per Capita GDP', color = 'Continent', title = 'Continent-Level Per Capita GDP') +
  scale_y_log10() +
  theme_minimal()

gdp_plot_log
```

# 6

```{r}
gdp_plot_lm <- ggplot(data = percapgdp, aes(x = year, y = percapgdp, color = continent)) +
  geom_point() +
  labs(x = 'Year', y = 'Per Capita GDP', color = 'Continent', title = 'Continent-Level Per Capita GDP') +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal()

gdp_plot_lm
```

```{r}
linear_model <- lm((percapgdp) ~ continent, data = percapgdp)

summary(linear_model)
```

1.  The mean per capita GDPs are increasing for all of the continents across from 1952-2007. Oceania maintained the highest per capita GDP.
2.  The coefficients are explaining the statistical significance compared to the per capita GDP of Africa. For every change in x, the y value changes by the value in est sd.
3. Africa, becuase it is in alphabetical order.
4. The coefficients represent the slope of the lines.
5. Yes, depending on the questions you are asking. This gives us a model to explain global scale per capita GDP, but does not have much granularity.We did not test the assumptions for linearity.
